---
title: "Homework 5"
author: "Jessica Murphy"
date: "February 26, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r load_packages, include = FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
```

## Overview

In this document, we will be working with Facebook data from the [UC Irvine Machine Learning Repository](https://archive.ics.uci.edu/ml/index.php). The dataset contains information about posts published on the Facebook page of a renowned cosmetics brand in 2014. More information can be found in [Facebook_metrics.txt](data/Facebook_metrics.txt).

## Data download

```{r data_download}
# Directory to store data in
data_dir <- "data"

# Create data directory if it doesn't exist
if(!dir.exists(data_dir)) {
  dir.create(data_dir)
}

# Files to save
data_zip <- paste(data_dir, "Facebook_metrics.zip", sep = "/")
data_file <- paste(data_dir, "dataset_Facebook.csv", sep = "/")

url_data <- "http://archive.ics.uci.edu/ml/machine-learning-databases/00368/Facebook_metrics.zip"

# Download and unzip file if it doesn't already exist 
if(!file.exists(data_zip)) {
    download.file(url_data, data_zip)
    unzip(data_zip, exdir = data_dir)
}
```

## Data reproducibility

```{r data_reproducibility}
library(tools)
md5 <- md5sum(data_file)
md5
file.info(data_file)
```

## Load data

```{r load_data}
# Notice the field separator in the data file is a semicolon
data <- read.csv(data_file, header = TRUE, sep = ";")
```

## Data processing

Here we add a column to the dataset with the day of the week spelled out using the Post.Weekday variable. We assume Sunday is 1.

```{r data_processing}
days_of_week <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
Day <- data$Post.Weekday

for (i in 1:7){
  Day[Day == i] <- days_of_week[[i]]
}

data$Day <- Day # add variable to data frame
```

## Exploratory analysis

Here we count the total number of each type of post.

```{r exploratory_analysis_1}
data %>% count(Type)
```

Here we create a ggplot boxplot of total interactions by type. We use a log10 scale for the y-axis.

```{r exploratory_analysis_2}
ggplot(data) +
  aes(x = Type, y = Total.Interactions) +
  geom_boxplot() +
  scale_y_log10()
```

Here we create another ggplot boxplot of total interactions by day. We again use a log10 scale for the y-axis and reorder the days for the x-axis.

```{r exploratory_analysis_3}
data$Day <- factor(data$Day, levels = days_of_week, ordered = TRUE)

ggplot(data) +
  aes(x = Day, y = Total.Interactions) +
  geom_boxplot() +
  scale_y_log10()
```

## Data analysis

Here we test to see if there is a significant difference in total interactions based on whether the posts were paid or not.

```{r data_analysis_1}
# Convert Paid into a categorical variable
data$Paid <- factor(data$Paid)
levels(data$Paid) <- c("not paid", "paid")

paid_interactions <- data %>% filter(Paid == "paid") %>% select(Total.Interactions)
unpaid_interactions <- data %>% filter(Paid == "not paid") %>% select(Total.Interactions)
t.test(paid_interactions, unpaid_interactions)
```

Based on the test results, there is not enough evidence to suggest there is a significant difference between the average total interactions of paid and unpaid posts using a significance level of 0.05. (p-value > 0.05)

Now we will create a boxplot of total interactions by whether the posts were paid or not to get a better look.

```{r data_analysis_2}
# Do not plot NA values
ggplot(subset(data, !is.na(Paid))) +
  aes(x = Paid, y = Total.Interactions) +
  geom_boxplot() +
  scale_y_log10() +
  theme(axis.title.x = element_blank())
```

## Software environment

```{r environment}
sessionInfo()
```