---
title: "Homework 5"
author: "Jessica Murphy"
date: "February 26, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this document, we will be working with Facebook data from the [UC Irvine Machine Learning Repository](https://archive.ics.uci.edu/ml/index.php). The dataset contains information about posts published on the Facebook page of a renowned cosmetics brand in 2014. More information can be found in the [Facebook_metrics](data/Facebook_metrics.txt) file.

## Data download

```{r data_download}
# Directory to store data in
data_dir <- "data"

# Create data directory if it doesn't exist
if(!dir.exists(data_dir)) {
  dir.create(data_dir)
}

# Files to save
data_zip <- paste(data_dir, "Facebook_metrics.zip", sep = "/")
data_file <- paste(data_dir, "dataset_Facebook.csv", sep = "/")

url_data <- "http://archive.ics.uci.edu/ml/machine-learning-databases/00368/Facebook_metrics.zip"

# Download and unzip file if it doesn't already exist 
if(!file.exists(data_zip)) {
    download.file(url_data, data_zip)
    unzip(data_zip, exdir = data_dir)
}
```

## Data reproducibility

```{r data_reproducibility}
library(tools)
md5 <- md5sum(data_file)
md5
file.info(data_file)
```

## Load data

```{r load_data}
data <- read.csv(data_file, header = TRUE, sep = ";")
```

## Data processing

```{r data_processing}
days_of_week <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
Day <- data$Post.Weekday

for (i in 1:7){
  Day[Day == i] <- days_of_week[[i]]
}

data$Day <- Day # add variable to data frame
```