---
title: "RNA-Seq Pre-processing"
author: "Jessica Murphy"
date: "March 15, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

### Overview

The following data was obtained from a 2x2 design using fly models. The two variables of interest are strain and tissue. The strains are white eyed (W) and sevenless (S) and the tissues are optic lope (O) and retina (R). There are four sample types: SO#, SR#, WO#, WR#.

### Raw Reads

The raw reads are located on Yampa under ```/BIOS6660/Homework6/rawReads/```. They are pair-ended reads with a read length of 150.

code: ```/home/murphjes/BIOS6660/Homework_6/Code/CountRawReads.sh```

### Trim Reads

The first 15 bases were trimmed off the reads using cudadapt v1..9.1, which gives us a read length of 135. The trimmed reads are located on Yampa under ```/home/murphjes/BIOS6660/Homework_6/TrimmedReads/```.

code: ```/home/murphjes/BIOS6660/Homework_6/Code/TrimReads.sh```

### Check for Quality

FastQC was used to evaluate the quality of the reads and check that the trimming went well. Ideally, we would like green checks on the Per Sequence Quality Scores and the Per Base Sequence Content. According to the reports, Sample A looked good, B was okay, and C and D were not as good. Since this data is a small subset of the original data, the quality is not as good. The reports are located on Yampa under ```/home/murphjes/BIOS6660/Homework_6/FastQC/```.

code: ```fastqc *.fastq.gz```

### Align Reads

The trimmed reads were then aligned to the dm6 genome using hisat2 v2.0.3. The reference files are located on Yampa under ```/BIOS6660/Homework6/indexFiles/``` and the aligned reads are located under ```/home/murphjes/BIOS6660/Homework_6/AlignedReads/```.

code: ```export PATH=/usr/local/bin/samtools-1.3:$PATH```
```hisat2 -x /BIOS6660/Homework6/indexFiles/dm6.hisat -1 /path/sample_R1_trimmed.fastq.gz -2 /path/sample_R2_trimmed.fastq.gz | samtools view -bS - > alignedSample.bam```

### Quantitate Reads

The trimmed files were also quantified to the Ensembl dm6 transcriptome using RSEM v1.2.31. The reference files are located on Yampa under ```/BIOS6660/Homework6/indexFiles/``` and the quantitation files are located under ```/home/murphjes/BIOS6660/Homework_6/Quantitation/```.

code: ```export PATH=/usr/local/bin/bowtie2:$PATH```
```rsem-calculate-expression -p 8 --time --seed 2020 --bowtie2 --paired-end --seed-length 20 /path/sample_R1_trimmed.fastq.gz /path/sample_R2_trimmed.fastq.gz /BIOS/Homework6/indexFiles/dm6.ensembl sampleName```

```{r, warning=FALSE, message=FALSE, error=FALSE, eval=FALSE, echo=FALSE}

#get the file names and sample names
 
wd = "C:/Users/jrain/AppData/Local/Packages/CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc/LocalState/rootfs/home/murphjes/BIOS6660/Homework_6/Quantitation/"

getGeneResults = function(a){
  b = a[grep(".gene",a)]
  return(b)
}
files =  paste(wd, getGeneResults(list.files(wd)), sep="")

getSample = function(a){
  b = sapply(strsplit(sapply(strsplit(a, ".genes", fixed=TRUE), "[[", 1), split="/", fixed=TRUE), "[[", 15)
  return(b)
}
samples = getSample(files)

#load in the data
for(i in 1:length(files)){
  x = read.table(file=files[i],sep="\t",header=TRUE)
  x = x[,c("gene_id","expected_count")]
  colnames(x)[2] = samples[i]
  if(files[i]!=files[1]) rsem = merge(x,rsem,by=c("gene_id"),all=TRUE)
  if(files[i]==files[1]) rsem = x
}

estCnts = rsem[,-1]
rownames(estCnts) = rsem$gene_id

counts = round(estCnts,0)
save(counts, file="C:/Users/jrain/AppData/Local/Packages/CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc/LocalState/rootfs/home/murphjes/BIOS6660/Homework_6/Data/RSEM_Counts.Rdata")
```
```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE}

load(file="C:/Users/jrain/AppData/Local/Packages/CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc/LocalState/rootfs/home/murphjes/BIOS6660/Homework_6/Data/RSEM_Counts.Rdata")
```

```{r, warning=FALSE, message=FALSE, error=FALSE, eval=FALSE, echo=FALSE}

wd = "C:/Users/jrain/AppData/Local/Packages/CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc/LocalState/rootfs/home/murphjes/BIOS6660/Homework_6/Quantitation/"

getTranscriptResults = function(a){
  b = a[grep(".isoform",a)]
  return(b)
}
files =  paste(wd, getTranscriptResults(list.files(wd)), sep="")

#load in the data
for(i in 1:length(files)){
  x = read.table(file=files[i],sep="\t",header=TRUE)
  x = x[,c("transcript_id","expected_count")]
  #colnames(x)[2] = files.v2[i, "sample"]
  if(files[i]!=files[1]) rsem = merge(x,rsem,by=c("transcript_id"),all=TRUE)
  if(files[i]==files[1]) rsem = x
}

estCnts = rsem[,-1]
rownames(estCnts) = rsem$transcript_id

transcript.counts = round(estCnts,0)
save(transcript.counts, file="C:/Users/jrain/AppData/Local/Packages/CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc/LocalState/rootfs/home/murphjes/BIOS6660/Homework_6/Data/RSEM_TranscriptCounts.Rdata")
```
```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE}
load(file="C:/Users/jrain/AppData/Local/Packages/CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc/LocalState/rootfs/home/murphjes/BIOS6660/Homework_6/Data/RSEM_TranscriptCounts.Rdata")
```

### Sample Counts

The following table shows the sum of reads for each sample at each point in the data processing.

```{r, warning=FALSE, message=FALSE, error=FALSE, echo=FALSE}

### raw reads
raw = read.table(file="C:/Users/jrain/AppData/Local/Packages/CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc/LocalState/rootfs/home/murphjes/BIOS6660/Homework_6/Data/RawReadsCounts.txt")
raw$sample = sapply(strsplit(sapply(strsplit(as.character(raw$V1), split="_R", fixed=TRUE), "[[", 1), split="rawReads_v2/", fixed=TRUE), "[[", 2)

raw2 = raw[,c("V3", "sample")]
colnames(raw2)[1] = "raw total Counts"
raw2 = raw2[!duplicated(raw2),]

### trimmed reads
trimmed = read.table(file="C:/Users/jrain/AppData/Local/Packages/CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc/LocalState/rootfs/home/murphjes/BIOS6660/Homework_6/Data/TrimmedReadsCount.txt")
trimmed$sample = sapply(strsplit(sapply(strsplit(as.character(trimmed$V1), split="_R", fixed=TRUE), "[[", 1), split="TrimmedReads/", fixed=TRUE), "[[", 2)

trimmed2 = trimmed[,c("V3", "sample")]
colnames(trimmed2)[1] = "trimmed total Counts"
trimmed2 = trimmed2[!duplicated(trimmed2),]

### rsem reads
rsem.sums = as.matrix(colSums(counts))
colnames(rsem.sums) = "RSEM total Counts"

m1 = merge(raw2, trimmed2, by="sample")
m2 = merge(m1, rsem.sums, by.x="sample", by.y=0)
m2$trimmPercent = (m2$`trimmed total Counts`/m2$raw)*100
m2$rsemPercent.fromTrimm = (m2$`RSEM total Counts`/m2$`trimmed total Counts`)*100

colnames(m2) = c("Sample", "Raw", "Trim", "RSEM", "Trim Percent", "RSEM Percent from Trim")
m2[,2] = prettyNum(m2[,2], big.mark=",")
m2[,3] = prettyNum(m2[,3], big.mark=",")
m2[,4] = prettyNum(m2[,4], big.mark=",")
m2[,5] = paste(round(m2[,5], 2), "%", sep="")
m2[,6] = paste(round(m2[,6], 2), "%", sep="")
```

`r kable(m2, align="c", row.names=FALSE)`