---
title: "Homework 4"
author: "Jessica Murphy"
date: "February 19, 2018"
output: html_document
---

### Introduction

We will be working with the [Traffic Counts](https://www.denvergov.org/opendata/dataset/city-and-county-of-denver-traffic-counts) dataset from the Denver Open Data Catalog. We demonstrate aspects of data management, code organization, and code quality.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r load_external, include = FALSE}
source("functions.R")
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
```


## Run the tests

Here we run our tests to check the implementations of our helper functions. If this document runs/knits without errors, that means the tests passed.

```{r run_tests}
source("tests.R")
```


### Load the dataset

```{r load_data}
data <- read.csv("../data/MASTER_TRAFFIC_COUNT.csv")
```


### Clean the data

```{r clean_data}
# Remove empty columns
data <- data %>% select(-starts_with("X"))
# Fix data types
data$Volume <- as.integer(data$Volume)
data$ADT <- as.integer(data$ADT)
data$AAWDT <- as.integer(data$AAWDT)
data$AAWET <- as.integer(data$AAWET)
# Standardize street names
data <- data %>% 
  mutate(Main.Location = standardize_street(Main.Location),
         Second.Location = standardize_street(Second.Location))
```


### Analyze the data

For this exercise we simply create one plot and save it to a file.

```{r }
plt <- plot_avg_wkday_traffic(data)
ggsave("../output/avg_wkday_traffic.png")
```


